---
title: "Motor_impairment_trial"
output:
  pdf_document: default
  html_document: default
---

#Setting base directory and loading libraries required for analysis
```{r}
baseDir=("/group/canc2/puumba/Data/InfiniumData/JeffCraig/VICS/Idat-SDF")
library(ggbio)
library(ExperimentHub)
library(missMethyl)
library(limma)
library(minfi)
library(IlluminaHumanMethylation450kanno.ilmn12.hg19)
library(ruv)
library(RColorBrewer)
library(matrixStats)
library(gplots)
library(WGCNA)
library(lumi)
library(ggplot2)
library(FlowSorted.CordBlood.450k)
library(tidyverse)
library(topconfects)

```
#Reading in annotation for 450k methylation arrays
```{r}
ann450 = getAnnotation(IlluminaHumanMethylation450kanno.ilmn12.hg19)
```
#Reading in the sample summary sheet (metadata)
```{r}
#Reading in the sample summary sheet (metadata)
targets_gen = read.metharray.sheet(baseDir, pattern = "motor_biom _ MABC_scores_mod.CSV")

```

#Reading array raw data and assigning sample names with array data
#targets$ID = paste(targets$Sample_Group,targets_gen$Sample_Name,sep=".")
```{r}
rgSet = read.metharray.exp(targets = targets_gen)
sampleNames(rgSet) = targets_gen$casenum

```
#Quality control
```{r}
detP = detectionP(rgSet)
qcReport(rgSet, sampNames=targets_gen$casenum, pdf="qc-report_motor_impairment.pdf")
pdf("mean_detection_motor_impairment.pdf",width=14)
par(mfrow=c(1,2))
cols=brewer.pal(4,"Set1")
barplot(apply(detP,2,mean),col=as.numeric(factor(targets_gen$casenum)),las=2,cex.names= 0.5, cex.axis=0.75,main="Mean detection p-values of probe signals",ylab="Mean detection p-value")
barplot(apply(detP,2,mean),col=as.numeric(factor(targets_gen$casenum)),las=2,cex.names= 0.5, cex.axis=0.75,ylim=c(0,0.010),main="Mean detection p-values of probe signals",ylab="Mean detection p-value")
dev.off()
```
#Preprocessing
```{r}
mset.raw = preprocessRaw(rgSet)
```
#Data exploration
```{r}
pdf("mds_plots_motor_impairment.pdf",width=14)
par(mfrow=c(1,2))
mdsPlot(mset.raw, sampGroups = targets_gen$ABCtotal_cutoff, sampNames=targets_gen$casename,legendPos="bottom",main="MABCtotal",pch=20)
mdsPlot(mset.raw, sampGroups = targets_gen$CP, sampNames=targets_gen$casename,legendPos="bottom",main="CP",pch=20)
mdsPlot(mset.raw, sampGroups = targets_gen$sex, sampNames=targets_gen$casename,legendPos="bottom",main="sex",pch=20)
mdsPlot(mset.raw, sampGroups = targets_gen$BPD_36_numeric, sampNames=targets_gen$casename,legendPos="bottom",main="BPD_36",pch=20)
mdsPlot(mset.raw, sampGroups = targets_gen$postnatal_steroids, sampNames=targets_gen$casename,legendPos="bottom",main="postnatal_steroids",pch=20)
mdsPlot(mset.raw, sampGroups = targets_gen$ga_week_cutoff, sampNames=targets_gen$casename,legendPos="bottom",main="gestational_age",pch=20)
dev.off()
```

#Normalisation
`
```{r}
mSetSw = preprocessSWAN(rgSet)
densityPlot(getBeta(mSetSw),main="SWAN")
```
```{r}
mSetSq = preprocessQuantile(rgSet)
pdf("normalisation_motor_impairment.pdf",width=14)
par(mfrow=c(1,3))
densityPlot(getBeta(mset.raw), sampGroups = targets_gen$casenum, main= "Raw", legend = FALSE)
densityPlot(getBeta(mSetSw), sampGroups = targets_gen$casenum,main="SWAN", legend = FALSE)
densityPlot(getBeta(mSetSq), sampGroups = targets_gen$casenum,main="SQN", legend = FALSE)
```

#Cell type composition analysis

```{r}

pdf("cell_type_motor_impairment21.pdf",width=14)
rgSet$Slide <- as.numeric(rgSet$Slide)
rgSet$Sex<- as.character(rgSet$sex)
rgSet$casenum<- as.character(rgSet$casenum)
cellCounts_new <- estimateCellCounts(rgSet, compositeCellType = "CordBlood", processMethod = "auto", probeSelect = "auto", cellTypes = c("CD8T","CD4T", "NK","nRBC","Bcell","Mono","Gran"), referencePlatform = c("IlluminaHumanMethylation450k"),returnAll = FALSE, meanPlot = TRUE)
write.csv(cellCounts_new,file="cell_counts.csv",row.names=TRUE)
MY.DATA<-read.csv("cell_counts.csv",  header=TRUE)
cellcounts2<-data.frame(MY.DATA)
colnames(cellcounts2)[1] <- "casenum"
view(cellcounts2)

targets_gen2<-merge(targets_gen, cellcounts2, by="casenum")
write.csv(targets_gen2,file="targets_gen2.csv",row.names=TRUE)
```

```{r}
#plot cell type composition by sample group
par(mfrow=c(1,1))
a = targets_gen2[targets_gen2$ABCtotal_cutoff == "1",]
b = targets_gen2[targets_gen2$ABCtotal_cutoff == "2",]
age.pal <- brewer.pal(8,"Set1")
boxplot(a, xlim=c(0, 18), xaxt="n",
        col=age.pal[1], main="", ylab="Cell type proportion")
boxplot(b, xaxt="n", add=TRUE, col=age.pal[2])
#axis(1, labels=colnames(a), tick=TRUE) 
#axis(1, labels=colnames(a), tick=TRUE)

age.pal <- brewer.pal(8,"Set1")
boxplot(a, at=0:5*3+ 1, xlim=c(0, 18), ylim=range(a,b, na.rm = TRUE), xaxt="n",
        col=age.pal[1], main="", ylab="Cell type proportion")
boxplot(b, at=0: 5*3+ 2, xaxt="n", add=TRUE, col=age.pal[2])
axis(1, at=0:5*3+ 1.5, labels=colnames(a), tick=TRUE) 

#legend("topleft", legend=c("Low","Moderate", "High"), fill=age.pal, cex=.7)

#wilcox.test(b[,"Gran"],c[,"Gran"], paired=FALSE)
#None of the p-values are less than 0.05 - separate word document in folder
```


#plot cell type composition by sample group
```{r}
pdf("cell_type_boxplot_motor_impairment.pdf",width=14)
boxplot(cellCounts_new,xlab="Cell type proportion",ylab="cell percentage")
```
#None of the p-values are less than 0.05 - separate word document in folder
#Filtering low quality probes
```{r}
keepProbes = rowSums(detP < 0.01) == ncol(detP)
mSetSqFlt = mSetSq[keepProbes,]
gmSetSqFlt = mapToGenome(mSetSqFlt)
```
#remove SNPs
```{r}
gmSetSqFlt = dropLociWithSnps(gmSetSqFlt, snps = c("CpG", "SBE"))
```

#Removing cross-reactive probes
```{r}
Xreact = read.csv(file="/group/canc2/puumba/Data/InfiniumData/JeffCraig/VICS/48639-non-specific-probes-Illumina450k.csv", stringsAsFactors=FALSE)
noXreact = !(featureNames(gmSetSqFlt) %in% Xreact$TargetID)
gmSetSqFlt = gmSetSqFlt[noXreact,]
```

#Removing probes on X and Y chromosomes
```{r}
autosomes = !(featureNames(gmSetSqFlt) %in% ann450$Name[ann450$chr %in% c("chrX","chrY")])
gmSetSqFlt = gmSetSqFlt[autosomes,]

```
#Relative log expression (RLE plot)

```{r}
mValsSq = getM(gmSetSqFlt)
medSq = apply(mValsSq, 1, median)
YSw = mValsSq - medSq
par(mfrow=c(1,2))
pdf("rle_plot_sep9.pdf")
boxplot(YSw,outline=FALSE,ylim=c(-1.5,1.5), ylab="Relative Log Methylation Value", cols=as.character(factor(targets_gen$ABCtotal_cutoff,)),xaxt="none")
title(xlab="Samples",cex=2, line=1)
dev.off()
```
```{r}
#MDS plots generation after filtering 
pdf("mds_plots_motor_impairment_normalisation.pdf",width=14)
par(mfrow=c(1,2))
mdsPlot(mValsSq, sampGroups = targets_gen$ABCtotal_cutoff, sampNames=targets_gen$casename,legendPos="bottom",main="MABCtotal",pch=20)
mdsPlot(mValsSq, sampGroups = targets_gen$CP, sampNames=targets_gen$casename,legendPos="bottom",main="CP",pch=20)
mdsPlot(mValsSq, sampGroups = targets_gen$sex, sampNames=targets_gen$casename,legendPos="bottom",main="sex",pch=20)
mdsPlot(mValsSq, sampGroups = targets_gen$BPD_36_numeric, sampNames=targets_gen$casename,legendPos="bottom",main="BPD_36",pch=20)
mdsPlot(mValsSq, sampGroups = targets_gen$postnatal_steroids, sampNames=targets_gen$casename,legendPos="bottom",main="postnatal_steroids",pch=20)
mdsPlot(mValsSq, sampGroups = targets_gen$ga_week_cutoff, sampNames=targets_gen$casename,legendPos="bottom",main="gestational_age",pch=20)
dev.off()
```

#MDS plots generation after filtering by dimension
```{r}

pal = brewer.pal(8, "Dark2")
mds1Sq = plotMDS(mValsSq, top=1000, gene.selection="common",dim.plot=c(1,2))
mds2Sq = plotMDS(mValsSq, top=1000, gene.selection="common",dim.plot=c(1,3))
mds3Sq = plotMDS(mValsSq, top=1000, gene.selection="common",dim.plot=c(2,3))
mds4Sq = plotMDS(mValsSq, top=1000, gene.selection="common",dim.plot=c(3,4))
pdf("MDS_plot_normalised_dimensions_motor_impairment.pdf",height=14,width=14)
par(mfrow=c(2,2))
plotMDS(mds1Sq, pch=19, col=pal[as.factor(targets_gen$ABCtotal_cutoff)])
plotMDS(mds2Sq, pch=19, col=pal[as.factor(targets_gen$ABCtotal_cutoff)])
plotMDS(mds3Sq, pch=19, col=pal[as.factor(targets_gen$ABCtotal_cutoff)])
plotMDS(mds4Sq, pch=19, col=pal[as.factor(targets_gen$ABCtotal_cutoff)])
plotMDS(mds1Sq, pch=19, col=pal[as.factor(targets_gen$postnatal_steroids)])
plotMDS(mds2Sq, pch=19, col=pal[as.factor(targets_gen$postnatal_steroids)])
plotMDS(mds3Sq, pch=19, col=pal[as.factor(targets_gen$postnatal_steroids)])
plotMDS(mds4Sq, pch=19, col=pal[as.factor(targets_gen$postnatal_steroids)])
plotMDS(mds1Sq, pch=19, col=pal[as.factor(targets_gen$ga_week_cutoff)])
plotMDS(mds2Sq, pch=19, col=pal[as.factor(targets_gen$ga_week_cutoff)])
plotMDS(mds3Sq, pch=19, col=pal[as.factor(targets_gen$ga_week_cutoff)])
plotMDS(mds4Sq, pch=19, col=pal[as.factor(targets_gen$ga_week_cutoff)])
plotMDS(mds1Sq, pch=19, col=pal[as.factor(targets_gen$BPD_36_numeric)])
plotMDS(mds2Sq, pch=19, col=pal[as.factor(targets_gen$BPD_36_numeric)])
plotMDS(mds3Sq, pch=19, col=pal[as.factor(targets_gen$BPD_36_numeric)])
plotMDS(mds4Sq, pch=19, col=pal[as.factor(targets_gen$BPD_36_numeric)])
plotMDS(mds1Sq, pch=19, col=pal[as.factor(targets_gen$sex)])
plotMDS(mds2Sq, pch=19, col=pal[as.factor(targets_gen$sex)])
plotMDS(mds3Sq, pch=19, col=pal[as.factor(targets_gen$sex)])
plotMDS(mds4Sq, pch=19, col=pal[as.factor(targets_gen$sex)])
plotMDS(mds1Sq, pch=19, col=pal[as.factor(targets_gen$CP)])
plotMDS(mds2Sq, pch=19, col=pal[as.factor(targets_gen$CP)])
plotMDS(mds3Sq, pch=19, col=pal[as.factor(targets_gen$CP)])
plotMDS(mds4Sq, pch=19, col=pal[as.factor(targets_gen$CP)])
plotMDS(mds1Sq, pch=19, col=pal[as.factor(targets_gen$Twin_Group)])
plotMDS(mds2Sq, pch=19, col=pal[as.factor(targets_gen$Twin_Group)])
plotMDS(mds3Sq, pch=19, col=pal[as.factor(targets_gen$Twin_Group)])
plotMDS(mds4Sq, pch=19, col=pal[as.factor(targets_gen$Twin_Group)])
dev.off()
```

#Principal Component Analysis (PCA)
```{r}
fit <- prcomp(t(mValsSq),center = TRUE, scale = TRUE,retx=TRUE)
loadings = fit$x
plot(fit,type="lines")
nGenes = nrow(mValsSq)
nSamples = ncol(mValsSq)
datTraits = targets_gen[,c(3,5,7,10,12,13,16,27,40,41,51,52)]
moduleTraitCor = cor(loadings[,1:6], datTraits, use = "p")
moduleTraitPvalue = corPvalueStudent(moduleTraitCor, nSamples)
par(cex=0.75, mar = c(6, 8.5, 3, 3))
textMatrix = paste(signif(moduleTraitCor, 2), "\n(", signif(moduleTraitPvalue, 1), ")", sep = "");
dim(textMatrix) = dim(moduleTraitCor)
pdf("PCA_heatmap_motor_impairment_blood.pdf",height=8,width=8)
labeledHeatmap(Matrix = t(moduleTraitCor), xLabels = colnames(loadings[,1:6]), yLabels = names(datTraits), colorLabels = FALSE, colors = blueWhiteRed(6), textMatrix = t(textMatrix), setStdMargins = FALSE, cex.text = 0.5, cex.lab.y = 0.6, zlim = c(-1,1), main = paste("PCA-trait relationships: Top principal components"))
dev.off()

```

Probe-wise differential methylation analysis

```{r}sex<-targets_gen$sex
ga<-targets_gen$ga_week
design <- model.matrix(~ABCtotal+sex+ga, data=targets_gen)
```
```{r}
targetsgen3<-targets_gen2[-which(targets_gen2$ABCtotal_cutoff==0),]
mDat =mValsSq[,-which(targets_gen2$ABCtotal_cutoff==0)]
bDat = ilogit2(mDat)
MABC<-targets_gen$ABCtotal_cutoff
design <- model.matrix(~ABCtotal_cutoff+sex+ga_week+nRBC+Bcell+Mono+NK+CD8T+
CD4T,  data=targetsgen3)
fit <- lmFit(mDat, design)
fit2 <- eBayes(fit)
summary(decideTests(fit2,p.value = 0.05))
top = topTable(fit2,coef="ABCtotal_cutoff",num=Inf)
ann_sub = 
  ann450[,c("chr","pos","strand","Name","Islands_Name","Relation_to_Island","UCSC_RefGene_Name","UCSC_RefGene_Group")]
output = merge(ann450,top,by.x="Name",by.y="row.names")
#output_full_epicanno = merge(ann_epic,top,by.x="Name",by.y="row.names")
write.csv(output[order(output$adj.P.Val),],file="Fullcohort_MABC_top_dmps_BIO2.csv",row.names=FALSE)
head(top)
confects<-limma_confects(fit,coef ="ABCtotal_cutoff", fdr = 0.05, step = 0.001, trend = FALSE,full = FALSE)
confects
fil <- tempfile("motor_impairment_trial.Rmd", fileext = ".rds")
saveRDS(confects, file = "fil", ascii = FALSE, version = NULL,compress = TRUE, refhook = NULL)



```

```{r}

mDat =mValsSq
design <- model.matrix(~CP+sex+ga_week +nRBC+Bcell+Mono+NK+CD8T+
CD4T, data=targets_gen2)
fit <- lmFit(mDat, design)
fit2 <- eBayes(fit)
summary(decideTests(fit2,p.value = 0.05))
top = topTable(fit2,coef="CP",num=Inf)
ann_sub = 
  ann450[,c("chr","pos","strand","Name","Islands_Name","Relation_to_Island","UCSC_RefGene_Name","UCSC_RefGene_Group")]
output = merge(ann450,top,by.x="Name",by.y="row.names")
#output_full_epicanno = merge(ann_epic,top,by.x="Name",by.y="row.names")
write.csv(output[order(output$adj.P.Val),],file="Fullcohort_CP_top_dmps_BIO2.csv",row.names=FALSE)
cp<-output[order(output$adj.P.Val)
 confects<-limma_confects(fit,coef ="CP", fdr = 0.05, step = 0.001, trend = FALSE,full = FALSE)
confects          
head(top)

```
```
preprocessNoob(rgSet, offset = 15, dyeCorr = TRUE, verbose = FALSE,
               dyeMethod=c("single", "reference"))

pheno <- pData(gset.fun)$casenum
designMatrix <- model.matrix(~ pheno)
dmrs_0.02 <- bumphunter(gset.funnorm, design = designMatrix, cutoff = 0.02, B=1000, type="Beta")
names(dmrs_0.02)
head(dmrs_0.02$table, n=20)
write.csv(dmrs_0.02$table, file = "bumphunter_1000.csv", row.names = TRUE)
```

```{r}
MOHANDASUP<-list()
MONHANDASDOWN<-
fgseaResup <- fgsea(MONHANDASUP,cp, maxSize=1000)
fgseaResdown <- fgsea(MONHANDASDOWN,cp, maxSize=1000)
```




```{r}
#scatter plots, histogram and boxplots
baseDir=("/group/canc2/puumba/Data/InfiniumData/jen. snow")
res <- read.csv("Fullcohort_CP_top_dmps.csv", header = TRUE)
ggplot(res, aes(x=P.Value))+ geom_histogram(binwidth=0.05,col="red", 
                 fill="green", 
                 alpha=.2)
cpg<-read.csv("boxplot_cpg.csv", header = TRUE)
ggplot(cpg, aes(x=CP, y=beta.cp)) + geom_boxplot() 
ggplot(cpg, aes(x=mabc, y=mABC.BETA)) + geom_boxplot(
ggplot(targets_gen2, aes(x=ABCtotal_cutoff, y=CD4T, )) + geom_boxplot()
ggplot(targets_gen2, aes(x=ABCtotal_cutoff, y=CD8T)) + geom_boxplot()
ggplot(targets_gen2, aes(x=ABCtotal_cutoff, y=Mono)) + geom_boxplot()
ggplot(targets_gen2, aes(x=ABCtotal_cutoff, y=Bcell)) + geom_boxplot()
ggplot(targets_gen2, aes(x=ABCtotal_cutoff, y=nRBC)) + geom_boxplot()
ggplot(targets_gen2, aes(x=ABCtotal_cutoff, y=NK)) + geom_boxplot()
dev.off()
```


```{r}

```


```{r}

library("DMRcate")
myannotation <- cpg.annotate("array", mDat, analysis.type="differential", design=design, coef=13, fdr = 0.1)
dmrcoutput <- dmrcate(myannotation, lambda=1000, C=2, pcutoff = 0.05)
results.ranges <- extractRanges(dmrcoutput, genome = "hg19") 
head(dmrcoutput$results)
length(dmrcoutput$results)
write.csv(dmrcoutput$results, file = "dmrcoutput.csv", row.names = TRUE)
coords <- strsplit2(dmrcoutput$results$coord[1],":")
chrom_dmrc <- coords[1]
start_dmrc <- as.numeric(strsplit2(coords[2],"-")[1])
end_dmrc <- as.numeric(strsplit2(coords[2],"-")[2]) 
minbase_dmrc <- start - (0.25*(end-start))
maxbase_dmrc <- end + (0.25*(end-start))
```
```{r}

```





```{r}

#To create stacked barplots
ggplot(targets_gen, aes(fill=sex, y=ABCtotal, x= ABCtotal)) + geom_bar(position="stack", stat="identity")

#to create scatter plot for two variables
ggplot(targets_gen2, aes(x=, y=ABCtotal)) + geom_point()

#to create split panel scatter plot 

#to create boxplots for cell types
ggplot(targets_gen, aes(x=celltype, y=percentage, fill=ABCtotal_cutoff)) +
  geom_boxplot() 

```{r}
#Gene Ontology analysis (gometh): top 1000 probes
#after sorting csv by pvalue
baseDir=("/group/canc2/puumba/Data/InfiniumData/jen. snow")
res <- read.csv("Fullcohort_MABC_top_dmps_BIO2.csv", header = TRUE)
sigCpGs_1k = res$rank[1:1000]
#total number
length(sigCpGs_1k)
sigCpGs_1k = as.character(sigCpGs_1k)
all = res
length(all)
#gometh()
par(mfrow=c(1,1))
gometh_1k <- gometh(sig.cpg=sigCpGs_1k, all.cpg=all, plot.bias=TRUE)
gometh_kegg <- gometh(sig.cpg = sigCpGs_1k, all.cpg = all, collection = "KEGG", prior.prob=TRUE)
topKEGG(gometh_kegg)
write.csv(gometh_1k, file = "GOterms_1k_cp.csv", row.names = TRUE)
write.csv(gometh_kegg, file = "GOmeth_kegg_cp.csv", row.names = TRUE)
#top GO terms
topGO(gometh_1k, ontology = "BP", number = 20L)
## Total number of significant GO categories at 5% FDR
```


```{r}
sum(gometh_kegg$FDR<0.5) 
```


```{r}

baseDir=("/group/canc2/puumba/Data/InfiniumData/jen. snow")
res <- read.csv("Fullcohort_CP_top_dmps_BIO2.csv", header = TRUE)
ggplot(res, aes(x=P.Value))+ geom_histogram(binwidth=0.05,col="red", 
                 fill="green", 
                 alpha=.2)
baseDir=("/group/canc2/puumba/Data/InfiniumData/jen. snow")
res <- read.csv("Fullcohort_MABC_top_dmps_BIO2.csv", header = TRUE)
ggplot(res, aes(x=P.Value))+ geom_histogram(binwidth=0.05,col="red", 
                 fill="green", 
                 alpha=.2)
baseDir=("/group/canc2/puumba/Data/InfiniumData/jen. snow")
res <- read.csv("Fullcohort_CP_top_dmps_CHAMP.csv", header = TRUE)
ggplot(res, aes(x=P.Value))+ geom_histogram(binwidth=0.05,col="red", 
                 fill="green", 
                 alpha=.2)
baseDir=("/group/canc2/puumba/Data/InfiniumData/jen. snow")
res <- read.csv("Fullcohort_MABC_top_dmps_champ.csv", header = TRUE)
ggplot(res, aes(x=P.Value))+ geom_histogram(binwidth=0.05,col="red", 
                 fill="green", 
                 alpha=.2)


```

```{r}
#Gene Ontology analysis (gometh): top 1000 probes
#after sorting csv by pvalue
baseDir=("/group/canc2/puumba/Data/InfiniumData/jen. snow")
res <- read.csv("Fullcohort_MABC_top_dmps_BIO2.csv", header = TRUE)
sigCpGs_1k = res$Name[1:1000]
#total number
length(sigCpGs_1k)
sigCpGs_1k = as.character(sigCpGs_1k)
all = res$Name
length(all)
#gometh()
par(mfrow=c(1,1))
gometh_1k <- gometh(sig.cpg=sigCpGs_1k, all.cpg=all, plot.bias=TRUE)
gometh_kegg <- gometh(sig.cpg = sigCpGs_1k, all.cpg = all, collection = "KEGG", prior.prob=TRUE)
topKEGG(gometh_kegg)
write.csv(gometh_1k, file = "GOterms_1k_MABC.csv", row.names = TRUE)
write.csv(gometh_kegg, file = "GOmeth_kegg_MABC.csv", row.names = TRUE)
#top GO terms
topGO(gometh_1k, ontology = "BP", number = 20L)
## Total number of significant GO categories at 5% FDR
sum(gometh_kegg$FDR<0.5) 
```










