---
title: "sarah's code"
output: html_document
---

```
```
```{r}
baseDir=("/group/canc2/puumba/Data/InfiniumData/jen. snow/project/Idat-SDF")
library(minfi)
targets_gen2 <- read.metharray.sheet(baseDir)
ann450 = getAnnotation(IlluminaHumanMethylation450kanno.ilmn12.hg19)
library(ChAMP)
myLoad <- champ.load(directory=baseDir,
                     filterXY=TRUE,
                     filterSNPs=TRUE,
                     arraytype="450K")
```

quality control
```{r}
champ.QC(beta=myLoad$beta,
         pheno=myLoad$pd$CP,
         resultsDir=baseDir,
         dendrogram=FALSE)
```

#Normalize
```
```
```{r}
myNorm <- champ.norm(arraytype="450K")

champ.QC(beta=myNorm,
         pheno=myLoad$pd$CP,
         resultsDir=baseDir,
         dendrogram=FALSE)
```

SVD
```{r}
library(dplyr, warn.conflicts = FALSE)
subpheno <- myLoad$pd %>%
  dplyr::select(ABCtotal_cutoff,
         ga_week_cutoff,
         sex,
         Slide,
         Array,
         CP,
        nRBC,
         Bcell,
         Mono,
        NK,
         CD8T,
         CD4T,
        Twin_Group)
subpheno2<-subpheno%>%rename(MABC=ABCtotal_cutoff, GA=ga_week_cutoff) 
champ.SVD(beta=myNorm,
          pd=subpheno,
          resultsDir="./CHAMP_SVDimages_Sarah/")

```
#In case you want to adjust for batch
```{r}
library(sva)
M <- logit2(myNorm)
modcombat <- model.matrix(~1, data=subpheno)
batch <- subpheno$Slide
myCombat <- ComBat(dat = M, batch = batch, mod = modcombat)
myCombat <- ilogit2(myCombat)

champ.SVD(beta=myCombat,
          pd=subpheno,
          resultsDir="./CHAMP_SVDimages_Sarah_batchcorrected/")
library(sva)
M <- logit2(myCombat)
modcombat <- model.matrix(~1, data=subpheno)
batch <- subpheno$Array
myCombat <- ComBat(dat = M, batch = batch, mod = modcombat)
myCombat <- ilogit2(myCombat)

champ.SVD(beta=myCombat,
          pd=subpheno,
          resultsDir="./CHAMP_SVDimages_Sarah_batchandpositioncorrected/")

```
Linear regression
subpheno=phenotype table
mycombat=DNA matrix
```{r}
subpheno2<- subpheno[-which(subpheno$ABCtotal_cutoff==0),]
mDat =myCombat[,-which(subpheno$ABCtotal_cutoff==0)]
design <- model.matrix(~ABCtotal_cutoff+sex+ga_week_cutoff+nRBC+ Bcell+Mono+NK+CD8T+
CD4T, data=subpheno2)
fit <- lmFit(mDat, design)
fit2 <- eBayes(fit)
summary(decideTests(fit2,p.value = 0.05))
top = topTable(fit2,coef="ABCtotal_cutoff",num=Inf)
ann_sub = 
  ann450[,c("chr","pos","strand","Name","Islands_Name","Relation_to_Island","UCSC_RefGene_Name","UCSC_RefGene_Group")]
output = merge(ann450,top,by.x="Name",by.y="row.names")
#output_full_epicanno = merge(ann_epic,top,by.x="Name",by.y="row.names")
write.csv(output[order(output$adj.P.Val),],file="Fullcohort_MABC_top_dmps_CHAMP.csv",row.names=FALSE)
head(top)
CHAMP<- limma_confects(fit,coef ="ABCtotal_cutoff", fdr = 0.05, step = 0.001, trend = FALSE,full = FALSE)
Mabc<-CHAMP[1:1000]
fil_CHAMP <- tempfile("CHAMP", fileext = ".rds")
saveRDS(CHAMP, file = "fil_CHAMP", ascii = FALSE, version = NULL,
        compress = TRUE, refhook = NULL)


```{r}
#Gene Ontology analysis (gometh): top 1000 probes
#after sorting csv by pvalue
baseDir=("/group/canc2/puumba/Data/InfiniumData/jen.snow/current.results")
res_CHAMP <- readRDS("fil_CHAMP", refhook = NULL)
sigCpGs_1k = res_CHAMP
#gometh()
par(mfrow=c(1,1))
gometh_1k <- gometh(sig.cpg=sigCpGs_1k, all.cpg=NULL, plot.bias=TRUE)
gometh_kegg <- gometh(sig.cpg = sigCpGs_1k, all.cpg = NULL, collection = "KEGG", prior.prob=TRUE)
topKEGG(gometh_kegg)
write.csv(gometh_1k, file = "GOterms_1k_mabc_CHAMP.csv", row.names = TRUE)
write.csv(gometh_kegg, file = "GOmeth_kegg__mabc_CHAMP.csv", row.names = TRUE)
#top GO terms
topGO(gometh_1k, ontology = "BP", number = 20L)
## Total number of significant GO categories at 5% FDR
sum(gometh_kegg$FDR<0.5) 
```
```{r}
library("DMRcate")
deign<-na.omit(design)
myannotation <- cpg.annotate("array", mDat, analysis.type="differential", design=deign, coef="ABCtotal_cutoff", fdr = 0.1)
dmrcoutput <- dmrcate(myannotation, lambda=1000, C=2, pcutoff = 0.05)
results.ranges <- extractRanges(dmrcoutput, genome = "hg19") 
head(dmrcoutput$results)
length(dmrcoutput$results)
write.csv(dmrcoutput$results, file = "dmrcoutput.csv", row.names = TRUE)
coords <- strsplit2(dmrcoutput$results$coord[1],":")
chrom_dmrc <- coords[1]
start_dmrc <- as.numeric(strsplit2(coords[2],"-")[1])
end_dmrc <- as.numeric(strsplit2(coords[2],"-")[2]) 



```

